Determinism, Complexity, and Predictability in Computer Performance

### 2014-08-04 model.tex
added mention of DVS fits
Liz final read through round I

\item The \emph{ARIMA} (auto-regressive integrated moving average)
  method, a common linear forecast strategy, built using the
  \emph{\arima} procedure \cite{autoARIMA}, and

@@

These four forecast methods were chosen to be a representative
sampling of standard prediction strategies, but they do not, of
course, cover that space exhaustively.  Our goal here is an empirical
assessment of the relationship between predictability and complexity,
not formal results about a ``best'' predictor for a given time series.
From a practitioner's standpoint, it would be useful to know \emph{ a
  priori} if a time series contains enough predictive structure to
make it worth spending the time and effort searching for a good
forecast method.  It would also be useful to know if a given method is
inadequate---that is, if an alternative method could do better and so
one should continue searching.  \alert{need to check that this is not
  redundant, given the other text in this section (and elsewhere in
  the paper).  it may be better to put this in the conclusion.}

@@

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear predictors.  Fitting such a model to a dataset involves
choosing values for the various free parameters in the autoregressive,
detrending, moving average, and filtering terms.  We employ the
automated fitting techniques described in~\cite{autoARIMA} to
accomplish this, producing what we will call an ``\arima model''in the
rest of this paper.  This procedure uses sophisticated methods---KPSS
unit-root tests~\cite{KPSSunit}, a customization of the Canova-Hansen
test~\cite{Canova1995}, and the Akaike information
criterion~\cite{akaike1974}, conditioned on the maximum likelihood of
the model fitted to the detrended data---to select good values for the
free parameters of the ARIMA model.

@@

Geometric structure in these kinds of plots is an indication of
structure in the information generation/transmission process that
produced the time series.  The dynamical systems community has
developed a number of methods that leverage this structure to generate
predictions (e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).
One of the most straightforward of these is the \emph{Lorenz method of
  analogues} (LMA), which is essentially nearest-neighbor prediction
in the embedded\footnote{Lorenz's original formulation used the full
  system state space;
%
%According to \cite{kantz97},
%
this method was first extended to embedded dynamics by
Pikovsky~\cite{pikovsky86-sov}, but is also related to the prediction
work of Sugihara \& May~\cite{sugihara90}}
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
prediction---provides accurate forecasts when the generating process
is a deterministic dynamical system.

Since LMA does not rest on an assumption of linearity (as ARIMA models
do), it can handle both linear and nonlinear processes.  If the
underlying generating process is nondeterministic, however, it can
perform poorly.  Figure~\ref{fig:embedding}(b), for instance, appears
to contain little structure, so LMA may not work well on this signal.
More structure appears to be present in Figure~\ref{fig:embedding}(c),
but this reconstruction also appears to contain some noise.  The
question as to how much structure is present in a reconstruction, and
how much of that structure can be captured and used by LMA, is apropos
of the central question treated in this paper.  It may be that \gcc
has some redundancy that LMA cannot exploit, or that the structure in
\svdfive is effectively obfuscated, from the standpoint of the LMA
method, by noise.  Without any knowledge of the generating process,
answers to these questions can only be derived from the data, with all
of the attendant problems (noise, sampling issues, and so on).  By
quantifying the balance between redundancy, predictive structure, and
entropy for these real-valued time series---as shown in
Section~\ref{sec:results}---we can begin to answer these questions in
an effective and practical manner.

@@

To study the relationship between predictability and complexity, we
use the four methods outlined above to generate predictions of all 120
traces described in Section~\ref{sec:methods}, then calculate the
error of the predictions with respect to the true continuations.
Specifically, we split each time series into two pieces: the first
90\%, referred to as the ``initial training" signal and denoted
$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
build the model, following the procedures described in the previous
section; that model is used to generate a prediction of the value of
$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
repeats $k$ times, out to the end of the observed time series.  This
``one step prediction'' process is not technically necessary in the
LMA method, whose ability to generate accurate predictions is limited
only by the positive Lyapunov exponents of the system.  However, the
performance of the other three methods used here will degrade severely
if the associated models are not periodically rebuilt.  In order to
make the comparison fair, we used an iterative one-step prediction
schema \emph{for all four methods}.  This has the slightly confusing
effect of causing the ``test'' signal to be used both to assess the
accuracy of each model and for periodic refitting.

@@

MASE scores for all 360 experiments are tabulated in the four middle
columns in Table~\ref{tab:error}.
%\begin{table*}
%\caption{Mean absolute scaled error (MASE) scores and weighted
%  permutation entropies for all eight processes studied in this paper.
%  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.
%  \alert{please reorder the columns from left to right to match the
%    order in which the methods were introduced, rather than in the
%    reverse of that order}}
%  \begin{center}
%  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
%  \hline\hline Signal & LMA MASE & \arima MASE & na\"{i}ve MASE & RW MASE & WPE \\
%  \hline
%
%  \col       & $0.050 \pm 0.002$ & $0.599 \pm 0.211$ & $0.571 \pm 0.002$  & $1.001 \pm 0.002$ & $0.513 \pm 0.003$ \\
%
%  \gcc       & $1.530 \pm 0.021$ & $1.837 \pm 0.016$ & $1.797 \pm 0.010$  & $1.138 \pm 0.011$ & $0.943 \pm 0.001$ \\
%
%  \svdone    & $0.827 \pm 0.076$ & $0.714 \pm 0.075$ & $2.676 \pm 4.328$  & $0.933 \pm 0.095$ & $0.957 \pm 0.016$ \\
%
%  \svdtwo    & $1.279 \pm 0.020$ & $2.163 \pm 0.027$ & $3.054 \pm 0.040$  & $1.125 \pm 0.012$ & $0.846 \pm 0.004$ \\
%
%  \svdthree  & $0.619 \pm 0.021$ & $0.713 \pm 0.010$ & $31.386 \pm 0.282$ & $0.707 \pm 0.009$ & $0.716 \pm 0.006$ \\
%
%  \svdfour   & $0.779 \pm 0.036$ & $0.979 \pm 0.032$ & $2.661 \pm 0.074$  & $1.034 \pm 0.035$ & $0.825 \pm 0.008$ \\
%
%  \svdfive   & $0.718 \pm 0.048$ & $2.370 \pm 0.051$ & $20.870 \pm 0.192$ & $1.001 \pm 0.047$ & $0.678 \pm 0.007$ \\
%
%  \svdsix    & $0.739 \pm 0.068$ & $1.438 \pm 0.061$ & $2.197 \pm 0.083$  & $1.060 \pm 0.055$ & $0.748 \pm 0.011$ \\
%
%  \hline\hline
%  \end{tabular*}
%  \end{center}
% \label{tab:error}
%  \end{table*}%
%  
 \begin{table*}
\caption{Mean absolute scaled error (MASE) scores and weighted
  permutation entropies for all eight processes studied in this paper.
  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.
  }
  \begin{center}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
  \hline\hline 
Signal & RW MASE & na\"{i}ve MASE & \arima MASE & LMA MASE & WPE \\
\hline
  \col       & $1.001 \pm 0.002$ & $0.571 \pm 0.002$  & $0.599 \pm 0.211$ & $0.050 \pm 0.002$ & $0.513 \pm 0.003$ \\
  \gcc       & $1.138 \pm 0.011$ & $1.797 \pm 0.010$  & $1.837 \pm 0.016$ & $1.530 \pm 0.021$ & $0.943 \pm 0.001$ \\
  \svdone    & $0.933 \pm 0.095$ & $2.676 \pm 4.328$  & $0.714 \pm 0.075$ & $0.827 \pm 0.076$ & $0.957 \pm 0.016$ \\
  \svdtwo    & $1.125 \pm 0.012$ & $3.054 \pm 0.040$  & $2.163 \pm 0.027$ & $1.279 \pm 0.020$ & $0.846 \pm 0.004$ \\
  \svdthree  & $0.707 \pm 0.009$ & $31.386 \pm 0.282$ & $0.713 \pm 0.010$ & $0.619 \pm 0.021$ & $0.716 \pm 0.006$ \\
  \svdfour   & $1.034 \pm 0.035$ & $2.661 \pm 0.074$  & $0.979 \pm 0.032$ & $0.779 \pm 0.036$ & $0.825 \pm 0.008$ \\
  \svdfive   & $1.001 \pm 0.047$ & $20.870 \pm 0.192$ & $2.370 \pm 0.051$ & $0.718 \pm 0.048$ & $0.678 \pm 0.007$ \\
  \svdsix    & $1.060 \pm 0.055$ & $2.197 \pm 0.083$  & $1.438 \pm 0.061$ & $0.739 \pm 0.068$ & $0.748 \pm 0.011$ \\  
    \hline\hline
  \end{tabular*}
  \end{center}
 \label{tab:error}
  \end{table*}%

In view of the discussion at the end of the previous paragraph, the
fact that the values in the second-from-left column are not
identically equal to 1.00 may be somewhat surprising.  This can happen
due to differences between in-sample and out-of-sample forecasting;
the last 10\% of \svdthree, for instance---the out-of-sample
signal---was more amenable to random-walk prediction, on the average,
than the first 90\%.  See Section~\ref{sec:results} for a deeper
discussion of this effect.

@@

In other cases, the correspondence between MASE score and the visual
appearance of these kinds of plots is not so clear cut.  The plots of
LMA predictions of \col and \svdfive both lie near the diagonal, for
instance, but the corresponding MASE scores were very different:
$0.050$ for \col and $0.718$ for \svdfive (resp., 20 and 1.4 times
better than random-walk forecasts of the same signals).  This is a
result of the normalization in the MASE score calculation.  Recall
from Section~\ref{sec:simple} that the random-walk method performs
especially poorly on signals that oscillate rapidly.  \col fits this
description, so the random-walk error on this signal---the denominator
of the four MASE scores in the first row of the Table---is
pathologically high, which skews those scores down.  Random-walk
prediction is very effective for the \svdfive signal, on the other
hand, so the denominator is small and the MASE scores are skewed
upwards.

>>>>

\item The \emph{ARIMA} (auto-regressive integrated moving average)
  method, a common linear forecast strategy, instatiated via the
  \emph{\arima} procedure \cite{autoARIMA}, and

@@

These four forecast methods were chosen to be a representative
sampling of standard prediction strategies, but they do not, of
course, cover that space exhaustively.  Our goal here is an empirical
assessment of the relationship between predictability and complexity,
not formal results about a ``best'' predictor for a given time series.
From a practitioner's standpoint, it would be useful to know \emph{ a
  priori} if a time series contains enough predictive structure to
make it worth spending the time and effort searching for a good
forecast method.  It would also be useful to know if a given method is
inadequate---that is, if an alternative method could do better and so
one should continue searching.  \alert{Liz to do: check that this is
  not redundant, given the other text in this section (and elsewhere
  in the paper).  it may be better to put this in the conclusion.}

@@

There is a vast amount of theory and literature regarding the
construction and use of models of this type; we refer the reader to
\cite{davislinearts} for an in-depth exploration.  For the purposes of
this paper, where the goal is to explore the relationship between
predictability and complexity across a broad array of forecast
strategies, seasonal ARIMA models are a good exemplar of the class of
linear predictors.  Fitting such a model to a dataset involves
choosing values for the various free parameters in the autoregressive,
detrending, moving average, and filtering terms.  We employ the
automated fitting techniques described in~\cite{autoARIMA} to
accomplish this, producing what we will call an ``\arima model'' in
the rest of this paper.  This procedure uses sophisticated
methods---KPSS unit-root tests~\cite{KPSSunit}, a customization of the
Canova-Hansen test~\cite{Canova1995}, and the Akaike information
criterion~\cite{akaike1974}, conditioned on the maximum likelihood of
the model fitted to the detrended data---to select good values for the
free parameters of the ARIMA model.

@@

Geometric structure in these kinds of plots is an indication of
structure in the information generation/transmission process that
produced the time series.  The dynamical systems community has
developed a number of methods that leverage this structure to generate
predictions (e.g.,~\cite{weigend-book,casdagli-eubank92,Smith199250}).
One of the most straightforward of these is the \emph{Lorenz method of
  analogues} (LMA), which is essentially nearest-neighbor prediction
in the embedded\footnote{Lorenz's original formulation used the full
  system state space;
%
%According to \cite{kantz97},
%
this method was first extended to embedded dynamics by
Pikovsky~\cite{pikovsky86-sov}, but is also related to the prediction
work of Sugihara \& May~\cite{sugihara90}}
space~\cite{lorenz-analogues}.  Even this simple algorithm---which
builds predictions by finding the nearest neighbor in the embedded
space of the given point, then taking that neighbor's path as the
prediction---provides remarkably accurate forecasts when the
generating process is a deterministic dynamical system.

Since LMA does not rest on an assumption of linearity (as ARIMA models
do), it can handle both linear and nonlinear processes.  If the
underlying generating process is nondeterministic, however, it can
perform poorly.  Figure~\ref{fig:embedding}(b), for instance, appears
to contain little structure, so one might not expect LMA to work well
on this signal.  More structure appears to be present in
Figure~\ref{fig:embedding}(c), but this reconstruction also appears to
contain some noise.  The question as to how much structure is present
in a reconstruction, and how much of that structure can be captured
and used by LMA, is apropos of the central question treated in this
paper.  It may be that \gcc has some redundancy that LMA cannot
exploit, or that the structure in \svdfive is effectively obfuscated,
from the standpoint of the LMA method, by noise.

Without any knowledge of the generating process, answers to these
questions can only be derived from the data, with all of the attendant
problems (noise, sampling issues, and so on).  By quantifying the
balance between redundancy, predictive structure, and entropy for
these real-valued time series---as shown in
Section~\ref{sec:results}---we can begin to answer these questions in
an effective and practical manner.

@@

To study the relationship between predictability and complexity, we
use the four methods outlined above to generate predictions of all 120
traces described in Section~\ref{sec:methods}, then calculate the
error of the predictions with respect to the true continuations.
Specifically, we split each time series into two pieces: the first
90\%, referred to here as the ``initial training" signal and denoted
$\{x_i\}_{i=1}^{n}$, and the last 10\%, known as the ``test" signal
$\{c_j\}_{j=n+1}^{k+n+1}$.  The initial training signal is used to
build the model, following the procedures described in the previous
section; that model is used to generate a prediction of the value of
$x_{n+1}$, which is then compared to the true continuation, $c_{n+1}$.
The model is then rebuilt using $\{x_i\}_{i=1}^{n+1}$ and the process
repeats $k$ times, out to the end of the observed time series.  This
``one step prediction'' process is not technically necessary in the
LMA method, whose ability to generate accurate predictions is limited
only by the positive Lyapunov exponents of the system.  However, the
performance of the other three methods used here will degrade severely
if the associated models are not periodically rebuilt.  In order to
make the comparison fair, we used an iterative one-step prediction
schema \emph{for all four methods}.  This has the slightly confusing
effect of causing the ``test'' signal to be used both to assess the
accuracy of each model and for periodic refitting.

@@

MASE scores for all 360 experiments are tabulated in the four middle
columns in Table~\ref{tab:error}.
%\begin{table*}
%\caption{Mean absolute scaled error (MASE) scores and weighted
%  permutation entropies for all eight processes studied in this paper.
%  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.
%  \alert{please reorder the columns from left to right to match the
%    order in which the methods were introduced, rather than in the
%    reverse of that order}}
%  \begin{center}
%  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
%  \hline\hline Signal & LMA MASE & \arima MASE & na\"{i}ve MASE & RW MASE & WPE \\
%  \hline
%
%  \col       & $0.050 \pm 0.002$ & $0.599 \pm 0.211$ & $0.571 \pm 0.002$  & $1.001 \pm 0.002$ & $0.513 \pm 0.003$ \\
%
%  \gcc       & $1.530 \pm 0.021$ & $1.837 \pm 0.016$ & $1.797 \pm 0.010$  & $1.138 \pm 0.011$ & $0.943 \pm 0.001$ \\
%
%  \svdone    & $0.827 \pm 0.076$ & $0.714 \pm 0.075$ & $2.676 \pm 4.328$  & $0.933 \pm 0.095$ & $0.957 \pm 0.016$ \\
%
%  \svdtwo    & $1.279 \pm 0.020$ & $2.163 \pm 0.027$ & $3.054 \pm 0.040$  & $1.125 \pm 0.012$ & $0.846 \pm 0.004$ \\
%
%  \svdthree  & $0.619 \pm 0.021$ & $0.713 \pm 0.010$ & $31.386 \pm 0.282$ & $0.707 \pm 0.009$ & $0.716 \pm 0.006$ \\
%
%  \svdfour   & $0.779 \pm 0.036$ & $0.979 \pm 0.032$ & $2.661 \pm 0.074$  & $1.034 \pm 0.035$ & $0.825 \pm 0.008$ \\
%
%  \svdfive   & $0.718 \pm 0.048$ & $2.370 \pm 0.051$ & $20.870 \pm 0.192$ & $1.001 \pm 0.047$ & $0.678 \pm 0.007$ \\
%
%  \svdsix    & $0.739 \pm 0.068$ & $1.438 \pm 0.061$ & $2.197 \pm 0.083$  & $1.060 \pm 0.055$ & $0.748 \pm 0.011$ \\
%
%  \hline\hline
%  \end{tabular*}
%  \end{center}
% \label{tab:error}
%  \end{table*}%
%  
 \begin{table*}
\caption{Mean absolute scaled error (MASE) scores and weighted
  permutation entropies for all eight processes studied in this paper.
  LMA $=$ Lorenz method of analogues; RW $=$ random-walk prediction.
  }
  \begin{center}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
  \hline\hline 
Signal & RW MASE & na\"{i}ve MASE & \arima MASE & LMA MASE & WPE \\
\hline
  \col       & $1.001 \pm 0.002$ & $0.571 \pm 0.002$  & $0.599 \pm 0.211$ & $0.050 \pm 0.002$ & $0.513 \pm 0.003$ \\
  \gcc       & $1.138 \pm 0.011$ & $1.797 \pm 0.010$  & $1.837 \pm 0.016$ & $1.530 \pm 0.021$ & $0.943 \pm 0.001$ \\
  \svdone    & $0.933 \pm 0.095$ & $2.676 \pm 4.328$  & $0.714 \pm 0.075$ & $0.827 \pm 0.076$ & $0.957 \pm 0.016$ \\
  \svdtwo    & $1.125 \pm 0.012$ & $3.054 \pm 0.040$  & $2.163 \pm 0.027$ & $1.279 \pm 0.020$ & $0.846 \pm 0.004$ \\
  \svdthree  & $0.707 \pm 0.009$ & $31.386 \pm 0.282$ & $0.713 \pm 0.010$ & $0.619 \pm 0.021$ & $0.716 \pm 0.006$ \\
  \svdfour   & $1.034 \pm 0.035$ & $2.661 \pm 0.074$  & $0.979 \pm 0.032$ & $0.779 \pm 0.036$ & $0.825 \pm 0.008$ \\
  \svdfive   & $1.001 \pm 0.047$ & $20.870 \pm 0.192$ & $2.370 \pm 0.051$ & $0.718 \pm 0.048$ & $0.678 \pm 0.007$ \\
  \svdsix    & $1.060 \pm 0.055$ & $2.197 \pm 0.083$  & $1.438 \pm 0.061$ & $0.739 \pm 0.068$ & $0.748 \pm 0.011$ \\  
    \hline\hline
  \end{tabular*}
  \end{center}
 \label{tab:error}
  \end{table*}%
In view of the discussion at the end of the previous paragraph, the
fact that the values in the second-from-left column are not
identically equal to 1.00 may be somewhat surprising.  This can happen
due to differences between in-sample and out-of-sample forecasting;
the last 10\% of \svdthree, for instance---the out-of-sample
signal---was more amenable to random-walk prediction, on the average,
than the first 90\%.  See Section~\ref{sec:results} for a deeper
discussion of this effect.

@@

In other cases, the correspondence between MASE score and the visual
appearance of these kinds of plots is not so clear cut.  The plots of
LMA predictions of \col and \svdfive both lie near the diagonal, for
instance, but the corresponding MASE scores are very different:
$0.050$ for \col and $0.718$ for \svdfive (resp., 20 and 1.4 times
better than random-walk forecasts of the same signals).  This is a
result of the normalization in the MASE score calculation.  Recall
from Section~\ref{sec:simple} that the random-walk method performs
especially poorly on signals that oscillate rapidly.  \col fits this
description, so the random-walk error on this signal---the denominator
of the four MASE scores in the first row of the Table---is
pathologically high, which skews those scores down.  Random-walk
prediction is very effective for the \svdfive signal, on the other
hand, so the denominator is small and the MASE scores are skewed
upwards.

### 2014-08-04 pecomplexity.tex
removing an alert and fixing a HEAD issue
editing the ARIMA fit paragraph
friday round I
Liz final read through round I

\title{
%Determinism, Complexity, and Predictability in Computer Performance\\
%Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
%\\
\alert{Quantifying Time-Series Predictability through Structural Complexity}
%\\
%Quantifying Predictability using Structural Complexity
%\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
%\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
}

@@

This paper provides insight into when, why, and how forecast
strategies fail when they are applied to complicated time series. We
conjecture that the inherent complexity of real-world time-series
data---which results from the dimension, nonlinearity, and
non-stationarity of the generating process, as well as from
measurement issues like noise, aggregation, and finite data
length---is both \emph{empirically} quantifiable and directly
correlated with predictability.  In particular, we argue that
\emph{redundancy} is an effective way to measure complexity and
predictive structure in an experimental time series and that
\emph{weighted permutation entropy} is an effective way to estimate
that redundancy.  To validate these conjectures, we study 120
different time-series data sets.  For each time series, we construct
predictions using a wide variety of forecast models, then compare the
accuracy of the predictions with the permutation entropy of that time
series.  We use the results to develop a model-free heuristic that can
help practitioners recognize when a particular prediction method is
not well matched to the task at hand: that is, when the time series
has more predictive structure than that method can capture and
exploit.

>>>>

\title{
%Determinism, Complexity, and Predictability in Computer Performance\\
%Quantifying structured complexity and its role in predictability: With applications to predicting computer performance.
%\\
% A model-free strategy for quantifying the predictability of
% experimental time-series data 
Model-free quantification of time-series predictability
% \alert{Quantifying Time-Series Predictability through Structural Complexity}
%\\
%Quantifying Predictability using Structural Complexity
%\\Decoding Predictability in Complicated Real-Valued Time Series: The balance between complexity and temporal structure
%\\Decoding Complicated Time Series: The balance between complexity and temporal structure and the implications for predictability
}

@@

\alert{This paper provides insight into when, why, and how forecast
  strategies fail when they are applied to complicated time series.}
We conjecture that the inherent complexity of real-world time-series
data---which results from the dimension, nonlinearity, and
non-stationarity of the generating process, as well as from
measurement issues like noise, aggregation, and finite data
length---is both \emph{empirically} quantifiable and directly
correlated with predictability.  In particular, we argue that
\emph{redundancy} is an effective way to measure complexity and
predictive structure in an experimental time series and that
\emph{weighted permutation entropy} is an effective way to estimate
that redundancy.  To validate these conjectures, we study 120
different time-series data sets.  For each time series, we construct
predictions using a wide variety of forecast models, then compare the
accuracy of the predictions with the permutation entropy of that time
series.  We use the results to develop a model-free heuristic that can
help practitioners recognize when a particular prediction method is
not well matched to the task at hand: that is, when the time series
has more predictive structure than that method can capture and
exploit.

### 2014-08-04 related.tex
Liz final read through round I

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most of the corresponding
solutions fall into two categories that we call model-based error
analysis and model-free information analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample
forecasting\footnote{The terms ``in sample'' and ``out of sample'' are
  used in different ways in the forecasting community.  Here, we
  distinguish those terms by the part of the time series that is the
  focus of the prediction: the observed data for the former and the
  unknown future for the latter.  In-sample forecasts---comparisons of
  predictions generated from \emph{part} of the observed time
  series---are useful for assessing model error and prediction
  horizons, among other things.}.
%
These distributions are then used as estimates of out-of-sample
forecast errors in those regions.  For example, Smith \emph{et al.}
make in-sample forecasts using ensembles around selected points in
order to predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series can exhibit varying levels of
local predictive uncertainty.  We expand on this finding later in this
paper with a time series that exhibits interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent predictability
of a time series and thus cannot be used to draw conclusions about
predictive structure that may be usable by other forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed \emph{post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to
calculate forecast errors and verify that they are normally distributed.
If they are not, that suggests that there is structure in the time
series that the model-building process was unable to capture and use.
The problem with this approach is lack of generality.
\label{page:normal-errors}
Normally distributed errors indicate that a model has captured the
structure in the data insofar as is possible, \emph{given the
  formulation of that particular model} (viz., the best possible
linear fit to a nonlinear dataset).  This gives no indication as to
whether another modeling strategy might do better.

@@

The model-based error analysis methods described in the previous three
paragraphs are based on specific assumptions about the underlying
generating process and knowledge about what will happen to the error
if those assumptions hold or fail.  Model-\emph{free} information
analysis moves away from those restrictions.  Our approach falls into
this class: we wish to measure the inherent complexity of an empirical
time series, then correlate that complexity with the predictive
accuracy of forecasts made using a number of different methods.

@@

More closely related is the work of Boffetta \emph{et
  al.}~\cite{boffetta02}, who investigate the scaling behavior of
finite-size Lyapunov exponents (FSLE) and $\epsilon$-entropy for a
wide variety of deterministic systems with known dynamics and additive
noise.  While the scaling of these measures acts as a general proxy
for predictability bounds, this approach differs from our work in a
number of fundamental ways.  First, Boffetta \emph{et al.} is a
theoretical study; they do not work with any real predictions.  We
focus on real-world time-series data, where one does not necessarily
have the ability to perturb or otherwise interact with the system of
interest, nor can one obtain or manufacture the (possibly large)
number of points that might be needed to estimate the
$\epsilon$-entropy for small $\epsilon$.  Second, we do not require
\emph{a priori} knowledge about the noise and its interaction with the
system.  Third, we tie information---in the form of the weighted
permutation entropy---directly to prediction error via calculated
values of a specific error metric.  Though FSLE and $\epsilon$-entropy
allow for the comparison of predictability between systems, they do
not directly provide an estimate of prediction error.  Finally, our
approach also holds for stochastic systems, where neither the FLSEs
nor their relationship to predictability is well defined.

>>>>

Quantifying predictability, which is sometimes called ``predicting
predictability,'' is not a new problem.  Most of the corresponding
solutions fall into two categories that we call model-based error
analysis and model-free information analysis.
%
%Actually this first class are both just analyzing model error distributions but some are local and some are global models but both are really just error
%
%predicting local predictive capacity (radial basis functions stuff, trying to predict error bounds on next forecast based on ensemble uncertainty) but does not aggregate tell you at what level the time series exhibits complexity only locally predictive structure, this actualy gets at the interesting point that different regions of a time series may exhibit differnt levels of complexity which we will illustrate with \svd
%
The first class focuses on errors produced by a fixed forecasting
schema.  This analysis can proceed locally or globally.  The local
version approximates error distributions for different regions of a
time-series model using local ensemble in-sample
forecasting\footnote{The terms ``in sample'' and ``out of sample'' are
  used in different ways in the forecasting community.  Here, we
  distinguish those terms by the part of the time series that is the
  focus of the prediction: the observed data for the former and the
  unknown future for the latter.  In-sample forecasts---comparisons of
  predictions generated from \emph{part} of the observed time
  series---are useful for assessing model error and prediction
  horizons, among other things.}.
%
These distributions are then used as estimates of out-of-sample
forecast errors in those regions.  For example, Smith \emph{et al.}
make in-sample forecasts using ensembles around selected points in
order to predict the local predictability of that time
series~\cite{Smith199250}.  This approach can be used to show that
different portions of a time series exhibit varying levels of
local predictive uncertainty.  We expand on this idea later in this
paper with a time series that exhibits interesting regime shifts.

Local model-based error analysis works quite well, but it only
approximates the \emph{local} predictive uncertainty \emph{in relation
  to a fixed model}.  It cannot quantify the inherent predictability
of a time series and thus cannot be used to draw conclusions about
predictive structure that may be usable by other forecast methods.
%
%Distrubtion of error. For many methods, if your error is not normally distributed this signals that there is stil more predictve structure to that could be used by for example a larger order ARMA process. But if error is normally distributed, this suggests that you have used up all the predictive structure that **that** model can use, this doesn't quantify if preidctuce structure exists that isn't being used by this process. For example, nonlinear structure which is ignored by a linear predictor.
%
Global model-based error analysis moves in this direction.  It uses
out-of-sample error distributions, computed \emph{post facto} from a
class of models, to determine which of those models was best.  After
building an autoregressive model, for example, it is common to
calculate forecast errors and verify that they are normally
distributed.  If they are not, that suggests that there is structure
in the time series that the model-building process was unable to
recognize, capture, and exploit.  The problem with this approach is
lack of generality.
\label{page:normal-errors}
Normally distributed errors indicate that a model has captured the
structure in the data insofar as is possible, \emph{given the
  formulation of that particular model} (viz., the best possible
linear fit to a nonlinear dataset).  This gives no indication as to
whether another modeling strategy might do better.

@@

The model-based error analysis methods described in the previous three
paragraphs are based on specific assumptions about the underlying
generating process and knowledge about what will happen to the error
if those assumptions hold or fail.  Model-\emph{free} information
analysis moves away from those restrictions.  Our approach falls into
this class: we wish to measure the inherent complexity of an empirical
time series, then study the correlation of that complexity with the
predictive accuracy of forecasts made using a number of different
methods.

@@

More closely related is the work of Boffetta \emph{et
  al.}~\cite{boffetta02}, who investigate the scaling behavior of
finite-size Lyapunov exponents (FSLE) and $\epsilon$-entropy for a
wide variety of deterministic systems with known dynamics and additive
noise.  While the scaling of these measures acts as a general proxy
for predictability bounds, this approach differs from our work in a
number of fundamental ways.  First, \cite{boffetta02} is a theoretical
study that does not involve any actual predictions.  We focus on
real-world time-series data, where one does not necessarily have the
ability to perturb or otherwise interact with the system of interest,
nor can one obtain or manufacture the (possibly large) number of
points that might be needed to estimate the $\epsilon$-entropy for
small $\epsilon$.  Second, we do not require \emph{a priori} knowledge
about the noise and its interaction with the system.  Third, we tie
information---in the form of the weighted permutation
entropy---directly to prediction error via calculated values of a
specific error metric.  Though FSLE and $\epsilon$-entropy allow for
the comparison of predictability between systems, they do not directly
provide an estimate of prediction error.  Finally, our approach also
holds for stochastic systems, where neither the FLSEs nor their
relationship to predictability is well defined.
